A Loop Quantm Gravity Inspired Action for the Bosonic String and Emergent Dimensions at Large Scales

### 2024-12-07 lqg-strings.tex
added EoMs and solution of the gauged sigma model to main file

First let us recall the form of the Polyakov action for the bosonic string \eqref{eqn:polyakov-action}:
\begin{equation*}
	S_{P} = -\frac{T}{2} \int \diff \tau \wedge \diff \sigma \sqrt{-g} g^{ab} \partial_a X^\mu \partial_b X^\nu \eta_{\mu\nu}.
\end{equation*}
If one was to ask a beginning graduate student as to how one can incorporate a gauge field in the above expression, the natural answer would be that we can promote the $X^\mu$ to vectors which live in the representation of some gauge group $\mc G$ and promote the ordinary derivative to a gauge covariant derivative \wrt a connection valued in $\mc G$. The resulting action would now have the form:
\begin{equation}\label{eqn:polyakov-gauged}
	S_{P} = -\frac{T}{2} \int \diff \tau \wedge \diff \sigma \sqrt{-g} g^{ab} D_a X^\mu D_b X^\nu \eta_{\mu\nu},
\end{equation}
where the covariant derivative $D_a$ is of the form:
\begin{equation}\label{eqn:covariant-deriv-v1}
	D_a X^\mu(\tau , \sigma) = \partial_a X^\mu + k \mc{A}_{a}^\mu{}_\nu X^\nu,
\end{equation}
where $\mc{A}_{a}^\mu{}_\nu$ is the gauge field and $k$ is the gauge coupling. A more experienced researcher would object saying that we know that the $X^\mu$ are spacetime co-ordinates and not vectors transforming under the gauge group $\mc G$. But we will see that this is not a problem. In fact, in order to be able to couple the string to an \emph{arbitrarily} curved geometry we will find that we \emph{require} the $X^\mu$ to transform as vectors under the action of some \emph{local} gauge field.

@@

It is useful to see the explicit form of a $\mf{iso}(3,1)$ gauge connection. The elements of the Lie algeba of $\mf{iso}(3,1)$ have the following form \cite{Wise2010MacDowell-Mansouri}
\begin{align}\label{eqn:poincare-lie-algeba}
	T^A{}_B & := \begin{pmatrix}
		             0          & b^1        & b^2        & b^3        & p^0/l \\
		             b^1        & 0          & j^3        & -j^2       & p^1/l \\
		             b^2        & -j^3       & 0          & j^1        & p^2/l \\
		             b^3        & j^2        & -j^1       & 0          & p^3/l \\
		             -\ep p^0/l & -\ep p^1/l & -\ep p^2/l & -\ep p^3/l & 0
	             \end{pmatrix} \nonumber \\
	        & = j^i J_i + b^i B_i + \frac{1}{l}p^a P_a,
\end{align}
where $J_i, B_i$ are the generators of rotations and boosts respectively. $P_a := (P_0, P_i)$ are the generators of translations. The parameter $\ep$ takes values in $\{-1,0,1\}$. Strictly speaking the above expression corresponds to an $\mf{iso}(3,1)$ Lie algebra generator only when $\ep = 0$. $l$ is a length scale which must be introduced so that the components of the ``vector'' $(p^0, p^i)$ can be viewed as components of a translation vector along the $3+1$ dimensional geometry $\mb{\Gamma}$. Similarly the boosts and rotations describe
where $J_i, B_i$ are the generators of rotations and boosts respectively. $P_a := (P_0, P_i)$ are the generators of translations. The parameter $\ep$ takes values in $\{-1,0,1\}$.

@@

Let us write down the gauged Polyakov action for the Poincaré lie algebra valued connection $\mc{A}_a$ whose components are given in \eqref{eqn:poincare-lie-algeba}:
\begin{equation}\label{eqn:poincare-polyakov}
	S'_{P} = -\frac{T}{2} \int \diff^2 x \sqrt{-g} g^{ab} \mc{D}_a X^A \mc{D}_b X^B G_{AB},
\end{equation}
where now the $X^A = \{X^\mu, X^r\}$ (as mentioned previously in \eqref{eqn:enlarged-Xs}) and $G_{AB}$ is the expanded metric on the $d+2$ enlarged bulk manifold $\mb{M}$ and the covariant derivative is of the form defined in \eqref{eqn:covariant-deriv-v1}, but only now acting on $X^A$ instead of on $X^\mu$:
\begin{equation}\label{eqn:covariant-deriv-v2}
	\mc{D}_a X^A(\tau , \sigma) = \partial_a X^A + k \mc{A}_{a}^A{}_B X^B.
\end{equation}
Now, we will expand out the action \eqref{eqn:poincare-polyakov} and express it in terms of the $X^\mu$ and $X^r$ separately. It is clear that the resulting expression will be sum of three terms, with the first term being the (gauged) Polyakov action for the string on $\mb{\Gamma}$, terms describing the coupling of the string on $\mb{\Gamma}$ with fluctuations along the extra dimensions $X^r$ and finally a Polyakov like term for the lone extra dimension. We can write this schematically as:
\begin{equation}\label{eqn:poincare-polyakov-decomp}
	S'_P[X^A] = S_P[X^\mu] + S_{int}[X^\mu, X^r] + S_P[X^r] ,
\end{equation}
or more explicitly as
\begin{multline} \label{eqn:poincare-polyakov-decomp-explicit}
	S' _{P} [X^{\mu} , X^{r}] = -\frac{T}{2} \int \diff ^2 x \sqrt{- g} g^{a b} \mc{D}_{a} X^{\mu} \mc{D}_{b} X^{\nu} \eta _{\mu \nu} - \\
	- T \int \diff ^2 x \sqrt{- g} g^{a b} \mc{D}_{a} X^{\mu} \mc{D}_{b} X^{r} G_{\mu r} - \\
	- \frac{T}{2} \int \diff ^2 x \sqrt{- g} g^{a b} \mc{D}_{a} X^{r} \mc{D}_{b} X^{r} G_{r r} ,
\end{multline}
where we can see a natural decomposition of the expanded bulk metric $G_{A B}$ as a $(d+1)$-dimensional tensor $\eta _{\mu \nu}$, a $(d+1)$-dimensional vector $V_{\mu} = G_{\mu r}$ and a scalar $V = G_{r r}$, akin to the ADM decomposition of the metric in standard GR literature.

@@

@@

In section \ref{sec:quantum-kg-string} we saw that our area-corrected model allows closed string spin-1 excitations, which is due to our solution not having distinct left and right sectors and the physical states not being invariant under rigid rotations of the string. Closed string excitations are usually related to the gravitational field and background spacetime itself, so what could the vector states of the closed string relate to? Well, in section \ref{sec:gauging-lt} we discussed a possible formulation of ST in terms of a Poincaré connection $\mc{A}_{\mu}$, which is an ingredient in describing the gravitational field in tetrad-connection \cite{Bodendorfer-LQG} formulation, and so this seems to be the natural candidate as to what the first level vector excitation of the closed string could be,
\begin{equation} \label{eqn:connection-string-vector-excitation}
	\mc{A}_{\mu} \sim \eta _{\mu \nu} (\Hat{a}^{\nu} _{1})^{\dagger} \ket{0} .
\end{equation}
Another candidate for what the vector excitations could relate to is the Ashtekar electric fields $\widetilde{E}^{\mu} _{I}$,
\begin{equation} \label{eqn:AEF-string-vector-excitation}
	\widetilde{E}^{\mu} _{I} \sim (\Hat{a}^{\mu} _{I})^{\dagger} \ket{0} .
\end{equation}

>>>>

First let us recall the form of the Polyakov action for the bosonic string \eqref{eqn:polyakov-action}:
\begin{equation*}
	S_{P} = -\frac{T}{2} \int \diff \tau \wedge \diff \sigma \sqrt{-g} g^{ab} \partial_a X^\mu \partial_b X^\nu \eta_{\mu\nu}.
\end{equation*}
If one was to ask a beginning graduate student as to how one can incorporate a gauge field in the above expression, the natural answer would be that we can promote the $X^\mu$ to vectors which live in the representation of some gauge group $\mc G$ and promote the ordinary derivative to a gauge covariant derivative \wrt a connection valued in $\mc G$. The resulting action would now have the form:
\begin{equation}\label{eqn:polyakov-gauged}
	S_{P} = -\frac{T}{2} \int \diff \tau \wedge \diff \sigma \sqrt{-g} g^{ab} D_a X^\mu D_b X^\nu \eta_{\mu\nu},
\end{equation}
where the covariant derivative $D_a$ is of the form:
\begin{equation}\label{eqn:covariant-deriv-v1}
	D_a X^\mu(\tau , \sigma) = \partial_a X^\mu + \mu \mc{A}_{a}^\mu{}_\nu X^\nu,
\end{equation}
where $\mc{A}_{a}^\mu{}_\nu$ is the gauge field and $k$ is the gauge coupling. A more experienced researcher would object saying that we know that the $X^\mu$ are spacetime co-ordinates and not vectors transforming under the gauge group $\mc G$. But we will see that this is not a problem. In fact, in order to be able to couple the string to an \emph{arbitrarily} curved geometry we will find that we \emph{require} the $X^\mu$ to transform as vectors under the action of some \emph{local} gauge field.

@@

It is useful to see the explicit form of a $\mf{iso}(3,1)$ gauge connection. The elements of the Lie algeba of $\mf{iso}(3,1)$ have the following form \cite{Wise2010MacDowell-Mansouri}
\begin{align}\label{eqn:poincare-lie-algeba}
	T^A{}_B & := \begin{pmatrix}
		             0          & b^1        & b^2        & b^3        & p^0/l \\
		             b^1        & 0          & j^3        & -j^2       & p^1/l \\
		             b^2        & -j^3       & 0          & j^1        & p^2/l \\
		             b^3        & j^2        & -j^1       & 0          & p^3/l \\
		             \ep p^0/l & -\ep p^1/l & -\ep p^2/l & -\ep p^3/l & 0
	             \end{pmatrix} \nonumber \\
	        & = j^i J_i + b^i B_i + \frac{1}{l}p^a P_a,
\end{align}
where $J_i, B_i$ are the generators of rotations and boosts respectively. $P_a := (P_0, P_i)$ are the generators of translations. The parameter $\ep$ takes values in $\{-1,0,1\}$. Strictly speaking the above expression corresponds to an $\mf{iso}(3,1)$ Lie algebra generator only when $\ep = 0$. $l$ is a length scale which must be introduced so that the components of the ``vector'' $(p^0, p^i)$ can be viewed as components of a translation vector along the $3+1$ dimensional geometry $\mb{\Gamma}$. Similarly the boosts and rotations describe
where $J_i, B_i$ are the generators of rotations and boosts respectively. $P_a := (P_0, P_i)$ are the generators of translations. The parameter $\ep$ takes values in $\{-1,0,1\}$.

@@

Let us write down the gauged Polyakov action for the Poincaré lie algebra valued connection $\mc{A}_a$ whose components are given in \eqref{eqn:poincare-lie-algeba}:
\begin{equation}\label{eqn:poincare-polyakov}
	S'_{P} = -\frac{T}{2} \int \diff^2 x \sqrt{-g} g^{ab} \mc{D}_a X^A \mc{D}_b X^B G_{AB},
\end{equation}
where now the $X^A = \{X^\mu, X^r\}$ (as mentioned previously in \eqref{eqn:enlarged-Xs}) and $G_{AB}$ is the expanded metric on the $d+2$ enlarged bulk manifold $\mb{M}$ and the covariant derivative is of the form defined in \eqref{eqn:covariant-deriv-v1}, but only now acting on $X^A$ instead of on $X^\mu$:
\begin{equation}\label{eqn:covariant-deriv-v2}
	\mc{D}_a X^A(\tau , \sigma) = \partial_a X^A + \mu \mc{A}_{a}^A{}_B X^B.
\end{equation}
Now, we will expand out the action \eqref{eqn:poincare-polyakov} and express it in terms of the $X^\mu$ and $X^r$ separately. It is clear that the resulting expression will be sum of three terms, with the first term being the (gauged) Polyakov action for the string on $\mb{\Gamma}$, terms describing the coupling of the string on $\mb{\Gamma}$ with fluctuations along the extra dimensions $X^r$ and finally a Polyakov like term for the lone extra dimension. We can write this schematically as:
\begin{equation}\label{eqn:poincare-polyakov-decomp}
	S'_P[X^A] = S_P[X^\mu] + S_{int}[X^\mu, X^r] + S_P[X^r] ,
\end{equation}
or more explicitly as
\begin{multline} \label{eqn:poincare-polyakov-decomp-explicit}
	S' _{P} [X^{\mu} , X^{r}] = -\frac{T}{2} \int \diff ^2 x \sqrt{- g} g^{a b} \mc{D}_{a} X^{\mu} \mc{D}_{b} X^{\nu} \eta _{\mu \nu} - \\
	- T \int \diff ^2 x \sqrt{- g} g^{a b} \mc{D}_{a} X^{\mu} \mc{D}_{b} X^{r} G_{\mu r} - \\
	- \frac{T}{2} \int \diff ^2 x \sqrt{- g} g^{a b} \mc{D}_{a} X^{r} \mc{D}_{b} X^{r} G_{r r} ,
\end{multline}
where we can see a natural decomposition of the expanded bulk metric $G_{A B}$ as a $(d+1)$-dimensional tensor $\eta _{\mu \nu}$, a $(d+1)$-dimensional vector $V_{\mu} = G_{\mu r}$ and a scalar $V = G_{r r}$, akin to the ADM decomposition of the metric in standard GR literature.

@@

\subsection{Gauged Polyakov Equations of Motion}

By varying the action \eqref{eqn:poincare-polyakov}, we get two sets of constraints from the variation of $g_{a b}$ and $A_{a} ^{A B}$, namely the vanishing of the worldsheet energy-momentum and spin-linear-momentum tensors $T_{a b}$ and $\sigma _{a} ^{A B}$
\begin{multline} \label{eqn:gauged-WS-EM-tensor}
	T_{a b} : = \mc{D}_{a} X^{A} \mc{D}_{b} X^{B} G_{A B} - \\
	- g_{a b} g^{c d} \mc{D}_{c} X^{A} \mc{D}_{d} X^{B} G_{A B} = 0
\end{multline}
\begin{equation}
	\sigma ^{A B} _{a} : = \mc{D}_{a} X^{[A} X^{B]} = 0 .
\end{equation}
The former allows us to solve for the worldsheet \textquotedblleft covariant metric" $g_{a b}$ as
\begin{equation} \label{eqn:gauged-WS-metric}
	g_{a b} = 2 f \mc{D}_{a} X^{A} \mc{D}_{b} X^{B} G_{A B}
\end{equation}
where
\begin{equation} \label{eqn:gauged-f-metric}
	\frac{1}{f} = g^{a b} \mc{D}_{a} X^{A} \mc{D}_{b} X^{B} G_{A B} ,
\end{equation}
while the latter should allow one to solve for $\mc{A}_{a}$ in terms of $X^{A}$.

The variation of the action \wrt $X^{A}$ results in the equation of motion of the embbeding fields
\begin{multline} \label{eqn:gauged-X-EoM}
	\mc{D}_{a} [\sqrt{- g} g^{a b} \mc{D}_{b} X^{B} G_{B C}] = \\
	= \sqrt{g} g^{a b} \mc{D}_{a} X^{A} \mc{D}_{b} X^{B} \partial _{C} G_{A B} ,
which in worldsheet \textquotedblleft covariant conformal gauge" $g^{a b} = \e ^{- \phi} \eta ^{a b}$ and for flat background geometry $G_{A B} = \eta ' _{A B}$ with
\begin{equation} \label{eqn:flat-bulk-metric}
	[\eta ' _{A B}] = \begin{bmatrix} [\eta _{\mu \nu}] & 0 \\ 0 & \epsilon \end{bmatrix}
simplifies to
\begin{equation}
	\eta ^{a b} \mc{D}_{a} \mc{D}_{b} X^{A} \eta ' _{A B} = 0 .
From here on, we'll work on the case of $\ep = - 1$ (adS background), for which the above equation can be expanded as
\begin{multline} \label{eqn:expanded-gauged-X-EoM}
	\eta ^{a b} \big( \del _{a} \del _{b} X^{A} + 2 \mu A^{A} _{a B} \del _{b} X^{B} + \\
	+ \mu \del _{a} A^{A} _{b B} X^{B} + \mu ^2 A^{A} _{a B} A^{B} _{b C} X^{C} \big) = 0 .
To produce the desired bulk geometry, the background connection can be taken as
\begin{equation} \label{eqn:adS-bulk-connection}
	\mc{A}_{\mu} = \begin{bmatrix} 0 & 0 & 0 & 0 & \delta ^{0} _{\mu} \\ 0 & 0 & 0 & 0 & \delta ^{1} _{\mu} \\ 0 & 0 & 0 & 0 & \delta ^{2} _{\mu} \\ 0 & 0 & 0 & 0 & \delta ^{3} _{\mu} \\ - \delta ^{0} _{\mu} & \delta ^{1} _{\mu} & \delta ^{2} _{\mu} & \delta ^{3} _{\mu} & 0 \end{bmatrix} ,
from which the pulledback connection $\mc{A}_{a} = \del _{a} X^{\mu} \mc{A}_{\mu}$ is thus
\begin{equation} \label{eqn:adS-WS-connection}
	\mc{A}_{a} = \begin{bmatrix} 0 & 0 & 0 & 0 & \del _{a} X^{0} \\ 0 & 0 & 0 & 0 & \del _{a} X^{1} \\ 0 & 0 & 0 & 0 & \del _{a} X^{2} \\ 0 & 0 & 0 & 0 & \del _{a} X^{3} \\ - \del _{a} X^{0} & \del _{a} X^{1} & \del _{a} X^{2} & \del _{a} X^{3} & 0 \end{bmatrix} ,
whereby imposing conformal gauge on the induced metric $\gamma _{a b} = \del _{a} X^{\mu} \del _{b} X^{\nu} \eta _{\mu \nu} = \e ^{\varphi} \eta _{a b}$ we can separate equations for the Minkowski sector $X^{\mu}$ and for the holographic dimension $X^{r}$:
\begin{multline} \label{eqn:Minkowski-sector-EoM}
	\eta ^{a b} \del _{a} \del _{b} X^{\mu} + \\
	+ \frac{\mu \eta ^{a b}}{1 + \mu X^{r}} \left( 2 \del _{a} X^{r} + k \del _{a} X^{\nu} X_{\nu} \right) \del _{b} X^{\mu} = 0
\begin{multline} \label{eqn:holographic-dim-EoM}
	\eta ^{a b} \del _{a} \del _{b} X^{r} + 2 \mu ^2 \e ^{\varphi} X^{r} + 4 \mu \e ^{\varphi} - \\
	- \frac{\mu ^2 \eta ^{a b}}{1 + \mu X^{r}} \left( 2 \del _{a} X^{r} + \mu \del _{a} X^{\nu} X_{\nu} \right) \del _{b} X^{\mu} X_{\mu} = 0 .
Assuming the perturbations on the string are small $|X^{A}| \ll 1$, we get a massive simplification
\begin{equation} \label{eqn:simplified-Minkowski-EoM}
	\eta ^{a b} \del _{a} \del _{b} X^{\mu} = 0
\begin{equation} \label{eqn:simplified-holographic-EoM}
	\eta ^{a b} \del _{a} \del _{b} X^{r} + 2 \mu ^2 \e ^{\varphi} X^{r} = - 4 \mu \e ^{\varphi} .
Thus, the Minkowski sector behaves like regular ST for small perturbations, and for the holographic dimension we can solve using a Green's function
\begin{equation} \label{eqn:Xr-GF}
	X^{r} (x) = - 4 \mu \int \diff ^2 x' G (x ; x') \e ^{\varphi (x')} ,
with the Green's function satisfying
\begin{equation} \label{eqn:Greens-function}
	(\del _{\sigma} ^2 - \del _{\tau} ^2 + 2 \mu ^2 \e ^{\varphi}) G (x ; x') = \delta ^2 (x , x') .
To find this Green's function, we must fix the conformal factor to be of the form $\e ^{\varphi} = f (\tau) - g (\sigma)$, thus allowing us to separate variables in the homogeneous case of eq \eqref{eqn:Greens-function} into
\begin{equation} \label{eqn:Green-X}
	X^{\prime \prime} - 2 \mu ^2 g (\sigma) X = - \lambda X
\begin{equation} \label{eqn:Green-T}
	\ddot{T} - 2 \mu ^2 f (\tau) T = - \lambda T .
These are Schrodinger's equations for potentials given by $\mu f (\tau)$ and $\mu g (\sigma)$, as such if we assume that $f$ and $g$ are harmonic oscillator potentials $f (\tau) = \omega ^2 \tau ^2 / 2$ and $g (\sigma) = k^2 \sigma ^2 / 2$ we can find known solutions for separation constant $\lambda = 2 \mu E = \mu k (2n + 1) = \mu \omega (2n + 1)$, so by the Spectral Theorem our separated spatial and temporal Green's functions are
\begin{equation} \label{eqn:X-function}
	X (\sigma , \sigma ') = \sum _{n \geq 0} - \frac{X_{n} (\sigma) X_{n} (\sigma ')}{\mu k (2 n + 1)}
\begin{equation} \label{eqn:T-function}
	T (\tau , \tau ') = \sum _{m \geq 0} - \frac{T_{m} (\tau) T_{m} (\tau ')}{\mu \omega (2 m + 1)} ,
where $X_{n}$ and $T_{m}$ are harmonic oscillator eigenfunctions. With this, the full Green's function $G = T.X$ and the solution for $X^{r}$ are thus
\begin{equation} \label{eqn:G-function}
	G (x ; x') = \sum _{m \geq 0} \sum _{n \geq 0} \frac{T_{m} (\tau) T_{m} (\tau ')}{\mu \omega (2 m + 1)} \frac{X_{n} (\sigma) X_{n} (\sigma ')}{\mu k (2 n + 1)}
\begin{multline} \label{eqn:Xr-solution}
	X^{r} (x) = - 4 \mu \sum _{m \geq 0} \sum _{n \geq 0} \frac{T_{m} (\tau)}{\mu \omega (2 m + 1)} \frac{X_{n} (\sigma)}{\mu k (2 n + 1)} \times \\
	\times \int \diff ^2 x' T_{m} (\tau ') X_{n} (\sigma ') \left( \frac{\omega ^2 \tau ^{\prime 2}}{2} - \frac{k^2 \sigma ^{\prime 2}}{2} \right) .

@@

In section \ref{sec:quantum-kg-string} we saw that our area-corrected model allows closed string spin-1 excitations, which is due to our solution not having distinct left and right sectors and the physical states not being invariant under rigid rotations of the string. Closed string excitations are usually related to the gravitational field and background spacetime itself, so what could the vector states of the closed string relate to? Well, in section \ref{sec:gauging-lt} we discussed a possible formulation of ST in terms of a Poincaré connection $\mc{A}_{\mu}$, which is an ingredient in describing the gravitational field in tetrad-connection \cite{Bodendorfer-LQG} formulation, and so this seems to be the natural candidate as to what the first level vector excitation of the closed string could be,
\begin{equation} \label{eqn:connection-string-vector-excitation}
	\mc{A}_{\mu} \sim \eta _{\mu \nu} (\Hat{a}^{\nu} _{1})^{\dagger} \ket{0} .
\end{equation}
Another candidate for what the vector excitations could relate to is the Ashtekar electric field $\widetilde{E}^{\mu} _{I}$,
\begin{equation} \label{eqn:AEF-string-vector-excitation}
	\widetilde{E}^{\mu} _{I} \sim (\Hat{a}^{\mu} _{I})^{\dagger} \ket{0} .
\end{equation}
